---
title: "Reproducible_Research_Project1"
author: "Wesley Loftie-Eaton"
date: "7/18/2019"
output: html_document
---


```{r, echo = FALSE}
# Load packages 
library(dplyr)
library(ggplot2)
```

### 1. Code for reading in the dataset and/or processing the data

```{r, cache = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#download.file(fileURL, destfile = "dat.zip", method = "curl")
unzip(zipfile = "dat.zip", exdir = "./data")

Download.date <- Sys.Date()

dat <- read.csv("data/activity.csv", header = TRUE)
```

* The data for this project was download on `r Download.date` from `r fileURL`.
* Rows with NAs are excluded

```{r}
# Print the first 20 rows
head(dat, 20)
```


### 2. Histogram of the total number of steps taken each day

```{r}
# Group by day and calculate the total steps for each day
Group.by.day <- group_by(dat, date)
Steps.day <- summarise(Group.by.day, Total = sum(steps))

# Use ggplot to plot a histogram of the total steps per day

plot.Steps.Day <- ggplot(Steps.day, aes(Total)) +
    labs(x = "Total steps per day", title = "Histogram of the total number of steps per day") +
    geom_histogram(bins = 20)

plot.Steps.Day
```


### 3. Mean and median number of steps taken each day

```{r}
summary(Steps.day)
```
The median and mean steps per day is 10765 and 10766, respectively


## 4. Time series plot of the average number of steps taken

```{r}
# Calculate the average number of steps for each interval across all days
Av.steps.interval <- aggregate(steps ~ interval, dat, mean)

# Plot a time series graph of the average steps for each interval
plot.Steps.interval <- ggplot(Av.steps.interval, aes(x = interval, y = steps)) +
    labs(x = "Interval", y = "Average steps per interval (all days)") +
    geom_line()

plot.Steps.interval
```


### 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r}
# Rank the the average steps in descending order
Most.steps <- with(Av.steps.interval, order(-steps))

# Extract the top-ranking interval
Most.steps2 <- Av.steps.interval[Most.steps[1], ]

Most.steps2
```

### 6. Code to describe and show a strategy for imputing missing data

```{r}
# Determine the number of NAs
num.NAs.before <- nrow(dat[!complete.cases(dat$steps), ])

# replace NAs with the average steps for that interval across all days
dat2 <- dat
for(i in 1:length(dat2[,1])){
    if (is.na(dat$steps[i])) {
        na.interval <- dat2$interval[i]
        av.steps <- Av.steps.interval[Av.steps.interval$interval == na.interval, ]
        dat2$steps[i] <- av.steps$steps
        }
}

# Confirm that there are no longer any NAs
num.NAs.after <- nrow(dat2[!complete.cases(dat2$steps), ])
```

There are `r num.NAs.before` NAs in the original dataset and `r num.NAs.after` NAs after they were removed. 

### 7. Histogram of the total number of steps taken each day after missing values are imputed

```{r}
# Using the dataset without missing values, regroup the data by day and sum the total steps per day
Group.by.day2 <- group_by(dat2, date)
Steps.day2 <- summarise(Group.by.day2, Total = sum(steps))

# Use ggplot to plot ahistogram of the total steps per day
plot.Steps.Day2 <- ggplot(Steps.day2, aes(Total)) +
    labs(x = "Total steps per day", title = "Histogram of the total number of steps per day after NA replacement") +
    geom_histogram(bins = 20)

plot.Steps.Day2
```


### 8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
# Catagorize days as weekday or weekend
dat2$day <- weekdays(as.Date(dat2$date))

Weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
Weekend <- c("Saturday", "Sunday")

dat.weekdays <- filter(dat2, day %in% Weekdays)
dat.weekend <- filter(dat2, day %in% Weekend)


# Calculate the average number of steps for each interval across all days
av.weekday.steps <- aggregate(steps ~ interval, dat.weekdays, mean)
av.weekday.steps$day <- "Weekday"

av.weekend.steps <- aggregate(steps ~ interval, dat.weekend, mean)
av.weekend.steps$day <- "Weekend"

Grouped.steps <- as.data.frame(rbind(av.weekday.steps, av.weekend.steps))

# Plot the data in two plots separated as weekday or weekend

plot.Steps.interval2 <- ggplot(Grouped.steps, aes(x = interval, y = steps)) +
    labs(x = "Interval", y = "Average steps per interval") +
    geom_line() +
    facet_grid(Grouped.steps$day ~ .)

plot.Steps.interval2
```